---
output:
  md_document:
    variant: markdown_github
---
# Demographics

### Authors: Stella Iacovides & Peter Kamerman

**Date: `r format(Sys.Date(), '%B %d, %Y')`**

***

## Load required packages and set chunk options
```{r LoadPackages, message = FALSE}
# Load packages
library(knitr)
library(pander)
library(ggplot2)
library(scales)
library(grid)
library(cowplot)
library(readr)
library(dplyr)
library(tidyr)

# knitr chunk options
opts_chunk$set(echo = FALSE,
               warning = FALSE,
               message = FALSE,
               fig.path = './figures/',
               fig.width = 7,
               fig.height = 6,
               dev = c('png', 'pdf'),
               tidy = TRUE, 
               tidy.opts = list(width.cutoff = 65))
```

## Load data
```{r import_demo, echo = TRUE}
demo <- read_csv('./data/demographics.csv', col_names = T)
```

## Quick look
```{r look_demo}
# dim(data)
# names(demo)
# head(demo)
# tail(demo)
glimpse(demo)
summary(demo)
```

## Summary plots
### Age
```{r age}
# table
pander(summary(demo$age_years), 
       caption = 'Five-number summary of age (years)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = age_years)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of age\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Age (years)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Weight
```{r weight}
# table
pander(summary(demo$weight_kg), 
       caption = 'Five-number summary of weight (kg)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = weight_kg)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of weight\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Weight (kg)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Height
```{r height}
# table
pander(summary(demo$height_m), 
       caption = 'Five-number summary of height (m)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = height_m)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of height\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Height (m)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Body mass index (BMI)
```{r bmi}
# table
pander(summary(demo$height_m), 
       caption = 'Five-number summary of body mass index (BMI; m/kg^2)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = bmi_kg.m)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of body mass index (BMI)\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Body mass index (kg/m^2)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Intensity of menstrual pain
```{r menstrualPain}
# table
pander(summary(demo$menstrual.pain_vas), 
       caption = 'Five-number summary of menstrual pain ratings (0-100mm VAS)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = menstrual.pain_vas)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of menstrual pain\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Pain intensity (0-100mm VAS)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### General Health Questionniare
```{r ghqScore}
# table
pander(summary(demo$ghq_score), 
       caption = 'Five-number summary of General Health Questionniare (GHQ) scores')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = ghq_score)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_hline(yintercept = 6, 
               colour = 'red',
               size = 1,
               linetype = 2) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of General Health Questionnaire\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'General Health Questionnaire score\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Pittsburg Sleep Quality Index (PSQI)
```{r psqiScore}
# table
pander(summary(demo$psi_score), 
       caption = 'Five-number summary of Pittsburg Sleep Quality Index (PSQI) scores')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = psi_score)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_hline(yintercept = 5, 
               colour = 'red',
               size = 1,
               linetype = 2) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of Pittsburg Sleep Quality Index\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Pittsburg Sleep Quality Index\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Age at menarche
```{r menarcheAge}
# table
pander(summary(demo$menarche.age_years), caption = 'Five-number summary of age at menarche (years)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = menarche.age_years)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of age of menarche\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Age of menarche (years)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Time since menarche
```{r timeMenarche}
# table
pander(summary(demo$menstruation_years), 
       caption = 'Five-number summary of time since menarche (years)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = menstruation_years)) +
   geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of time since menarche\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Time since menarche (years)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Average menstrual cycle length
```{r cycleLength}
# table
pander(summary(demo$cycle.length_days), 
       caption = 'Five-number summary of menstrual cycle length (days)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = cycle.length_days)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of menstrual cycle length\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Length of menstrual cycle (days)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

### Duration of menstrual period
```{r periodLength}
# table
pander(summary(demo$menses.length_days), 
       caption = 'Five-number summary of duration of menstrual period (days)')

# plot
ggplot(data = demo, 
       aes(x = factor('Participants'), 
           y = menses.length_days)) +
    geom_boxplot(outlier.size = 3,
                 fatten = 3) +
    geom_jitter(shape = 21,
                size = 5,
                fill = 'orange',
                colour = '#D55E00',
                position = position_jitter(height = 0)) +
    labs(title  = 'Box-n-Whisker plot of duration of menstrual period\n(scatterplot of individual data: orange circles)',
         x = '',
         y = 'Menstrual period (days)\n') +
    theme_cowplot() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = 18),
          axis.title = element_text(size =18),
          plot.title = element_text(size = 18))
```

## Session information
```{r sessionInfo}
sessionInfo()
```