---
output:
  md_document:
    variant: markdown_github
---
# Regression

### Authors: Stella Iacovides & Peter Kamerman

**Date: `r format(Sys.Date(), '%B %d, %Y')`**

***

## Load required packages and set chunk options
```{r LoadPackages, message = FALSE, warning = FALSE}
# Load packages
library(ggplot2)
library(scales)
library(grid)
library(cowplot)
library(PMCMR) # pairwise Friedman posthoc tests
library(metafor) # arcsine transform
library(nlme) # mixed effect 'lme'
library(gamlss) # GAM using beta distribution
library(lsr) # Cohen's D
library(knitr)
library(pander)
library(readr)
library(dplyr)
library(tidyr)
library(boot) # Bootstrap 95% CI

# Load palette
cb8.categorical <- c("#0072B2", "#D55E00", "#009E73", "#F0E442", "#56B4E9", "#E69F00", "#CC79A7", "#999999")

# set seed
set.seed(123)

# knitr chunk options
opts_chunk$set(echo = TRUE,
               warning = FALSE,
               message = FALSE,
               fig.path = './figures/',
               fig.width = 7,
               fig.height = 7,
               dev = c('png', 'pdf'),
               tidy = TRUE, 
               tidy.opts = list(width.cutoff = 65))
```

## Load data
```{r import_pain}
data <- read_csv('./data/regression.csv')
```

## Quick look
```{r look_pain}
# dim(data)
# names(data)
# head(data)
# tail(data)
glimpse(data)
# summary(data)
```

## Process data
```{r process_pain}
# Convert id and period to factors
ischaemia <- data %>%
    mutate(id = factor(id), 
           period = factor(period))

# Convert to long format table
ischaemia_analysis <- ischaemia %>%
    group_by(id, period) %>%
    gather(key = time, 
           value = score, 
           3:13) %>%
    mutate(time = as.numeric(as.character(time)),
           score = score/100)

# Process data for singlePanel plot
ischaemia_plot <- ischaemia %>%
    group_by(id, period) %>%
    gather(key = time, value = score, 3:13) %>%
    mutate(time = as.numeric(as.character(time)))

## Process data for facetPanel plot 
ischaemia_plot.f <- ischaemia_plot %>%
    mutate(period = factor(period,
                           levels = c('baseline',
                                      'fragmentation1',
                                      'fragmentation2'),
                           labels = c('Baseline night',
                                      'Fragmentation night 1',
                                      'Fragmentation night 2')))
```

## Data analysis
### Plot the data
```{r singlePlot}
# Plot a single panel plot of all data
ggplot(ischaemia_plot, 
       aes(x = time,
           y = score,
           colour = period,
           fill = period)) +
    geom_smooth(method = 'loess', alpha = 0.5, size = 2) +
    labs(x = '\nTime (minutes)',
         y = 'Pain intensity (0-100mm VAS)\n',
         title = 'Pain intensity for 10 minutes after ischaemia\n') +
    scale_y_continuous(limits = c(0, 100),
                       expand = c(0,0)) +
    scale_x_continuous(limits = c(0, 10),
                       expand = c(0, 0),
                       breaks = c(0, 2, 4, 6, 8, 10),
                       labels = c(0, 2, 4, 6, 8, 10)) +
    scale_fill_manual(name = 'Intervention',
                      labels = c('Baseline night',
                                 'Fragmentation night 1',
                                 'Fragmentation night 2'),
                      values = cb8.categorical) +
    scale_colour_manual(name = 'Intervention',
                      labels = c('Baseline night',
                                 'Fragmentation night 1',
                                 'Fragmentation night 2'),
                      values = cb8.categorical) +
    theme_cowplot() +
    theme(legend.position = c(0.78, 0.12),
          legend.text = element_text(size = 18),
          legend.title = element_text(size = 18),
          legend.background = element_rect(fill = 'gray90'),
          axis.text = element_text(size = 18),
          axis.title = element_text(size = 18),
          plot.title = element_text(size = 18))
```

```{r facetPlot}
# Multi-facet plot of all data
ggplot(ischaemia_plot.f, 
       aes(x = time,
           y = score,
           colour = period,
           fill = period)) +
    geom_smooth(method = 'loess', alpha = 0.5, size = 2) +
    geom_jitter() +
    facet_grid(period~.) +
    labs(x = '\nTime (minutes)',
         y = 'Pain intensity (0-100mm VAS)\n',
         title = 'Facet plot: pain intensity for 10 minutes after ischaemia\n') +
    scale_y_continuous(limits = c(00, 100),
                       expand = c(0,0)) +
    scale_x_continuous(limits = c(0, 10),
                       expand = c(0, 0),
                       breaks = c(0, 2, 4, 6, 8, 10),
                       labels = c(0, 2, 4, 6, 8, 10)) +
    scale_fill_manual(name = 'Intervention',
                      labels = c('Baseline night',
                                 'Fragmentation night 1',
                                 'Fragmentation night 2'),
                      values = cb8.categorical) +
    scale_colour_manual(name = 'Intervention',
                      labels = c('Baseline night',
                                 'Fragmentation night 1',
                                 'Fragmentation night 2'),
                      values = cb8.categorical) +
    theme_cowplot() +
    theme(legend.position = 'none',
          panel.margin.y = unit(1, 'lines'),
          axis.text = element_text(size = 18),
          axis.title = element_text(size = 18),
          plot.title = element_text(size = 18))
```

## Regression analysis
### Raw VAS scores
```{r rawVAS}
# Check data distribution
histDist(ischaemia_analysis$score,
         main = 'Density plot of raw VAS scores\n(expressed as a proportion)',
         xlab = 'VAS',
         ylab = 'Denisty')

## Linear mixed model of raw VAS score (mod1)
## with variable slope and intercept
### Null model
mod1.null <- lme(score ~ 1,
                 random = ~ time | id,
                 data = ischaemia_analysis,
                 method = 'ML')
summary(mod1.null)
### Basic model (no intervention)
mod1.basic <- lme(score ~ time,
                 random = ~ time | id,
                 data = ischaemia_analysis,
                 method = 'ML')
summary(mod1.basic)
### Full model (time + intervention)
mod1.full <- lme(score ~ time + period,
                 random = ~ time | id,
                 data = ischaemia_analysis,
                 method = 'ML')
summary(mod1.full)

# Is the basic model fit better than that of the null model?
anova(mod1.null, mod1.basic)
# Is the basic model fit better than that of the null model?
anova(mod1.null, mod1.full)
# Is the basic model fit better than that of the basic model?
anova(mod1.basic, mod1.full)

# Diagnostic plots
## Residuals vs fitted
plot(fitted(mod1.full), residuals(mod1.full), main = 'Residuals vs fitted')
abline(h = 0, col = 'red', lwd = 2)

## Residuals vs mixed model index
plot(residuals(mod1.full), main = 'Residuals vs mixed model index')
abline(h = 0, col = 'red', lwd = 2)

## Distribution of residuals
histDist(residuals(mod1.full), 
         main = 'Residual density distribution',
         ylab = 'density')

## QQ plot of residuals
qqnorm(residuals(mod1.full))
qqline(residuals(mod1.full), col = 'red', lwd = 2)
```

### Arcsine transformed VAS scores
```{r arcsineVAS}
# Create new column with VAS data transformed using arcsine square root transform
ischaemia_analysis <- ischaemia_analysis %>%
    mutate(score.arcsine = transf.arcsin(score))

# Check data distribution
histDist(ischaemia_analysis$score.arcsine,
         main = 'Density plot of arcsine transformed VAS scores',
         xlab = 'VAS',
         ylab = 'Denisty')

## Linear mixed model of arcsine transformed VAS scores (mod2)
## with variable slope and intercept
### Null model
mod2.null <- lme(score.arcsine ~ 1,
                 random = ~ time | id,
                 data = ischaemia_analysis,
                 method = 'ML')
summary(mod2.null)
### Basic model (no intervention)
mod2.basic <- lme(score.arcsine ~ time,
                 random = ~ time | id,
                 data = ischaemia_analysis,
                 method = 'ML')
summary(mod2.basic)
### Full model (time and intervention)
mod2.full <- lme(score.arcsine ~ time + period,
                 random = ~ time | id,
                 data = ischaemia_analysis,
                 method = 'ML')
summary(mod2.full)

# Is the basic model fit better than that of the null model?
anova(mod2.null, mod2.basic)
# Is the full model fit better than that of the null model?
anova(mod2.null, mod2.full)
# Is the full model fit better than that of the basic model?
anova(mod2.basic, mod2.full)

# Diagnostic plots
## Residuals vs fitted
plot(fitted(mod2.full), residuals(mod2.full), main = 'Residuals vs fitted')
abline(h = 0, col = 'red', lwd = 2)

## Residuals vs mixed model index
plot(residuals(mod2.full), main = 'Residuals vs mixed model index')
abline(h = 0, col = 'red', lwd = 2)

## Distribution of residuals
histDist(residuals(mod2.full), 
         main = 'Residual density distribution',
         ylab = 'density')

## QQ plot of residuals
qqnorm(residuals(mod2.full))
qqline(residuals(mod2.full), col = 'red', lwd = 2)
```

### GAM using raw VAS scores, with beta-regression link
```{r gamVAS}
## Mixed model general additive model based on an expanded 
## (includes 0 and 1) beta distribution (mod3)
## with variable slope and intercept
### Null model
mod3.null <- gamlss(score ~ 1 + re(random = ~time|id), 
                     data = ischaemia_analysis,
                     family = BEINF)
summary(mod3.null)
### Basic model (no intervention)
mod3.basic <- gamlss(score ~ time + re(random = ~time|id), 
                     data = ischaemia_analysis,
                     family = BEINF)
summary(mod3.basic)
### Full model (time and intervention)
mod3.full <- gamlss(score ~ time + period + re(random = ~time|id), 
                     data = ischaemia_analysis,
                     family = BEINF)
summary(mod3.full)

## To determine whether there is a difference between fragmentation nights
## 1 and 2, reorder 'period' factor so that fragmentation2 is the 
## reference factor
ischaemia.analysis_frag2ref <- ischaemia_analysis %>%
    mutate(period = factor(period, 
                           levels = c('fragmentation2',
                                      'fragmentation1',
                                      'baseline'),
                           labels = c('fragmentation2',
                                      'fragmentation1',
                                      'baseline')))
mod3.full_frag2ref <- gamlss(score ~ time + period + re(random = ~time|id), 
                     data = ischaemia.analysis_frag2ref,
                     family = BEINF)
summary(mod3.full_frag2ref)

# Choose model
## Inspect generalised AIC 
GAIC(mod3.null, mod3.basic, mod3.full, mod3.full_frag2ref)
## Likelihood Ratio test for nested GAMLSS models
### Is the basic model fit better than that of the null model?
LR.test(mod3.basic, mod3.null)
### Is the full model fit better than that of the null model?
LR.test(mod3.null, mod3.full)
### Is the full model fit better than that of the basic model?
LR.test(mod3.basic, mod3.full)

# Diagnostic plots
## Residuals vs fitted
plot(fitted(mod3.full), residuals(mod3.full), main = 'Residuals vs fitted')
abline(h = 0, col = 'red', lwd = 2)

## Residuals vs mixed model index
plot(residuals(mod3.full), main = 'Residuals vs mixed model index')
abline(h = 0, col = 'red', lwd = 2)

## Distribution of residuals
histDist(residuals(mod3.full), 
         main = 'Residual density distribution',
         ylab = 'density')

## QQ plot of residuals
qqnorm(residuals(mod3.full))
qqline(residuals(mod3.full), col = 'red', lwd = 2)
```

## Differences in pain intensity at t = 0
```{r deltaT0}
# Determine whether the is a difference in pain intensity
# at the end of the ischaemia (t = 0) between the three nights.

# Filter by t = 0 scores, and arcsine transform the data
ischaemia_transf <- ischaemia_analysis %>%
    filter(time == 0) %>%
    mutate(score = transf.arcsin(score))
# Analyse with one-way anova
summary(aov(score ~ period + Error(id), 
            data = ischaemia_transf))
```

## Effect size
### Change in pain score within each night
```{r effectSize_1}
# Calculate a bootstrap mean and confidence interval for the 
# change in pain score over time for each night 
# (baseline, and fragmentation nights 1 and 2)

# Extract t = 0 and t = 10 minutes
delta <- ischaemia %>%
    mutate(change = `0` - `10`) %>%
    select(id, period, change) %>%
    spread(period, change)

# Boot function
boot_func <- function(data, d){
    median(data[d])
} 
# 95% CI
## Baseline
baseline_out <- boot(delta$baseline, 
                     boot_func, R = 10000)
baseline_ci <- boot.ci(baseline_out, type = 'norm')
cat(paste0("# Baseline night\n\nBootstrap mean (95% CI) change in pain intensity (0-100mm VAS) over a\nten-minute period following removal of a torniquet:\n", 
       round(baseline_out$t0), "mm",
       " (", round(baseline_ci$normal[2]), 
       " to ", round(baseline_ci$normal[3]), ")"))
plot(baseline_out)
## fragmentation 1
frag1_out <- boot(delta$fragmentation1, 
                  boot_func, R = 10000)
frag1_ci <- boot.ci(frag1_out, type = 'norm')
cat(paste0("# One night of sleep fragmentation\n\nBootstrap mean (95% CI) change in pain intensity (0-100mm VAS) over a\nten-minute period following removal of a torniquet:\n", 
       round(frag1_out$t0), "mm",
       " (", round(frag1_ci$normal[2]), " to ",
       round(frag1_ci$normal[3]), ")"))
plot(frag1_out)
## fragmentation 2
frag2_out <- boot(delta$fragmentation2, 
                  boot_func, R = 10000)
frag2_ci <- boot.ci(frag2_out, type = 'norm')
cat(paste0("# Two nights of sleep fragmentation\n\nBootstrap mean (95% CI) change in pain intensity (0-100mm VAS) over a\nten-minute period following removal of a torniquet:\n", 
       round(frag2_out$t0), "mm",
       " (", round(frag2_ci$normal[2]), " to ",
       round(frag2_ci$normal[3]), ")"))
plot(frag2_out)
```

### Cohen's D
```{r effectSize_2}
# Calculate Cohen's D after transforming the data (arcsine)
cohens_d <- ischaemia %>%
    select(id, period, `0`, `10`) %>%
    mutate(`0` = transf.arcsin(`0`/100), 
           `10` = transf.arcsin(`10`/100)) %>%
    mutate(change = `0` - `10`) %>%
    select(id, period, change)

# Baseline vs fragmentation night 1
BvF1 <- cohens_d %>%
    filter(period == 'baseline' | period == 'fragmentation1') %>%
    spread(period, change) %>%
    transmute(baseline, fragmentation = fragmentation1)
cohensD(BvF1$baseline, BvF1$fragmentation, method = 'paired')

# Baseline vs fragmentation night 2
BvF2 <- cohens_d %>%
    filter(period == 'baseline' | period == 'fragmentation2') %>%
    spread(period, change) %>%
    transmute(baseline, fragmentation = fragmentation2)
cohensD(BvF2$baseline, BvF2$fragmentation, method = 'paired')
```

## Session information
```{r sessionInfo}
sessionInfo()
```
